---
title: "2D Finite Difference Model"
author: "Claire Harrigan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include = FALSE}
# This RMarkdown file creates an html file of a finite difference model.

# Clear all from workspace/environment.
  rm(list = ls())

# Set universal settings for all code chunks.
  knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

```{r ModelDimensions}
# Define size of model domain.
  x.size <- 100000  # [m] 
  z.size <- 100000  # [m] 

# Set resolution (number of nodes).  
  x.num <- 40 
  z.num <- 40 
  
# Define stepsize.
  dx <- x.size / (x.num - 1)  # [m] 
  dz <- z.size / (z.num - 1)  # [m]  

# Create vectors of node locations.
  x <- seq(from = 0, to = x.size, by = dx)  # [m]
  z <- seq(from = 0, to = z.size, by = dz)  # [m]
```

```{r Constants}
# Define material properties, initialize arrays.
  k <- matrix(data = 3.0,      # thermal conductivity, [W/(m*K)]
              nrow = z.num,
              ncol = x.num)  
  A <- matrix(data = 2.2e-8,   # heat production, [W/(m^3)] (1e-6)
              nrow = z.num,
              ncol = x.num)  
  rho <- matrix(data = 3300,   # density, [kg/(m^3)] (2800)
                nrow = z.num,
                ncol = x.num)
  Cp <- matrix(data = 1293,    # heat capacity, [J/(kg*K)] (1000)
               nrow = z.num,
               ncol = x.num)
  Kappa <- k / (rho * Cp)      # [(m^2)/s]
  Kappa.min <- min(Kappa)      # [(m^2)/s] # Kappa.min <- 1e-6   # (Buchanan)
  U <- matrix(data = 0,        # uplift, [m/s]
              nrow = z.num,
              ncol = x.num)
  H <- matrix(data = 0,        # latent heat production, value? units?
              nrow = z.num,
              ncol = x.num)  
  # Input geothermal gradient in K/km.
  a.Kkm <- 20  # [K/km]
  a <- a.Kkm / 1000  # [K/m]
                   
# Define time step.
  dt.max <- (dx ^ 2) / (6 * Kappa.min)  # [s] 
  dt.yr <- dt.max * 3.17098e-8 # [yr]
  dt <- dt.max
```

```{r InitializeTDistribution_plume}
# Set up an initial condition of a circular plume at an elevated temperature.
  
# Initialize a matrix for initial temperature conditions.
  T.0 <- matrix(data = 900,
                nrow = z.num,
                ncol = x.num)

# Set the top row to be air temperature.
  # Choose a value for the air temperature.
  T.surface <- 25  # [°C]
  # Assign the first row of the initial matrix to be the air temperature.
  T.0[1, ] <- T.surface + 273.15

# Make a geotherm for the initialization matrix.
  for (j in 2:dim(T.0)[1]) {
    T.0[j, ] <- T.0[1, 1] + a * x[j]
  }
  
# Instantanteously emplace a melt sheet and set material properties of the melt
# sheet.
  # Set dimensions of melt sheet.
  meltsheetdiameter <- 20000  # [m]
  meltsheetthickness <- 3000  # [m]
  # If the melt sheet is centered in the x direction in the model, find the 
  # position in meters along the x axis.
  xstart.melt <- mean(x) - (meltsheetdiameter / 2)  # x position, [m]
  xend.melt <- mean(x) + (meltsheetdiameter / 2)  # x position, [m]
  # Find the index in the vector that corresponds to the position above.
  xstart.melt.index <- which(abs(x - xstart.melt) == min(abs(x - xstart.melt)))
  xend.melt.index <- which(abs(x - xend.melt) == min(abs(x - xend.melt)))
  # Make the melt sheet start on the second row of the matrix, below the air
  # layer.
  zstart.melt.index <- 2 
  # If the melt sheet is meltsheetthickness and starts at the second row of the
  # model, find the index that corresponds to the bottom of the melt sheet.
  zend.melt <- z[2] + meltsheetthickness
  zend.melt.index <- which(abs(z - zend.melt) == min(abs(z - zend.melt)))
  
  # Set temperature of melt sheet.
  T.meltsheet <- 3500  # [°C]
  
  # Emplace melt sheet. Set material properties of melt sheet.
  for (i in zstart.melt.index:zend.melt.index) {
    for (j in xstart.melt.index:xend.melt.index) {
      T.0[i, j] <- T.meltsheet + 273.15
      # k[i, j]     <- 3.0    # thermal conductivity, [W/(m*K)]
      # A[i, j]     <- 1e-6   # heat production, [W/(m^3)]
      # rho[i, j]   <- 3300   # density, [kg/(m^3)]
      # Cp[i, j]    <- 1293   # heat capacity, [J/(kg*K)]
      # Kappa[i, j] <- k[i, j] / (rho[i, j] * Cp[i, j])  # [(m^2)/s]
    }
  }

# Clean up.
  rm(i, j)
```

### Initial Condtions  
  
**Model Parameters:**  
dt = `r round((dt * 3.17098e-8 / 1000000), digits = 3)` Myr)
dx = `r round(dx, digits = 1)` m  
dz = `r round(dz, digits = 1)` m  

**Material Properties:**  
Heat production, A = `r mean(A)` W/(m^3)   
Thermal conductivity, k = `r mean(k)` W/(m K)  
Density, rho = `r mean(rho)` kg/(m^3)  
Heat capcity, Cp = `r mean(Cp)` J/(kg K)  
Uplift, U = `r mean(U)` m/s  

**Initial perterbation:**  
circular intrusion with `r (r.plume / 1000)` km radius at an elevated temperature (1700°C)  

```{r PlotInitialConditions_heatmap}
# Plot initial conditions.

# Load packages plotly and Rcolorbrewer for plotting.
  library(plotly)
  library(RColorBrewer)

# Make function for plotting.
  PlotlyMeltSheet <- function(temperatures) {
    plot_ly(x = ~(x / 1000),
            y = ~(z / 1000), 
            z = ~(temperatures - 273.15),
            type = "heatmap",
            colors = rev(brewer.pal(9,"RdYlBu"))) %>%
      add_trace(y = ((z[[2]] - z[[1]])/ 2 / 1000),
                type = "scatter",
                mode = "lines",
                line = list(color = "black")) %>%
      colorbar(title = "Temperature (°C)",
               limits = c(0, 3500))  %>%
      layout(xaxis = list(title = "Distance (km)"),
             yaxis = list(title = "Depth (km)",
                          autorange = "reversed"))
  }

# Plot initial condtions as a heatmap.
  PlotlyMeltSheet(T.0)

  # plot_ly(x = ~(x / 1000),
  #         y = ~(z / 1000), 
  #         z = ~(T.0 - 273.15),
  #         type = "heatmap",
  #         colors = rev(brewer.pal(9,"RdYlBu"))) %>%
  #   add_trace(y = ((z[[2]] - z[[1]])/ 2 / 1000),
  #             type = "scatter",
  #             mode = "lines",
  #             line = list(color = "black")) %>%
  #   colorbar(title = "Temperature (°C)",
  #            limits = c(0, 3500))  %>%
  # layout(xaxis = list(title = "Distance (km)"),
  #        yaxis = list(title = "Depth (km)",
  #                     autorange = "reversed"))
  # 
  
```

```{r Model_SetIterations}
# Set a number of time steps.
  n <- 10  # 100 iterations is about 5 myr

# Initialize a list as a container for model returns at different times.
  T.n  <- list()
  
# Place T.0 in the first position of the T.n list.
  T.n[[1]] <- T.0

# Solve the finite difference equation for n time steps.
  
  for (t in 1:n) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n)` iterations

Maximum temperature change between model run n = `r n` (time = `r round((n * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap}
# Plot model results after n iterations.
  PlotlyMeltSheet(T.n[[length(T.n)]])
```


```{r Model_SetIterations_2}
# Set a number of time steps.
  m <- 100  # 100 iterations is about 5 myr

# Solve the finite difference equation for n time steps.
  
  for (t in n:(n + m)) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n + m)` iterations

Maximum temperature change between model run n = `r (n + m)` (time = `r round(((n + m) * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_2}
# Plot model results after n iterations.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.n[[length(T.n)]] - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```

```{r Model_SetIterations_3}
# Set a number of time steps.
  o <- 100  # 100 iterations is about 5 myr

# Solve the finite difference equation for n time steps.
  
  for (t in n:(n + m + o)) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n + m + o)` iterations

Maximum temperature change between model run n = `r (n + m + o)` (time = `r round(((n + m + o) * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_3}
# Plot model results after n iterations.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.n[[length(T.n)]] - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```
