---
title: "Melt Sheet 2D"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include = FALSE}
# This RMarkdown file creates an html file of a finite difference model.

# Clear all from workspace/environment.
  rm(list = ls())

# Set universal settings for all code chunks.
  knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

```{r ModelDimensions}
# Define size of model domain.
  # x.size <- 160000  # [m] 
  # z.size <- 15000  # [m] 
  x.size <- 100000  # [m] 
  z.size <- 100000  # [m] 
  
# Set resolution (number of nodes).  
  # x.num <- 160 
  # z.num <- 15 
  x.num <- 100 
  z.num <- 100
  
# Define stepsize.
  dx <- x.size / (x.num - 1)  # [m] 
  dz <- z.size / (z.num - 1)  # [m]  

# Create vectors of node locations.
  x <- seq(from = 0, to = x.size, by = dx)  # [m]
  z <- seq(from = 0, to = z.size, by = dz)  # [m]
```

```{r Constants}
# Define material properties, initialize arrays.
  k <- matrix(data = 2.5,      # thermal conductivity, [W/(m*K)]
              nrow = z.num,
              ncol = x.num)  
  A <- matrix(data = 0,   # heat production, [W/(m^3)] (1e-6)
              nrow = z.num,
              ncol = x.num)  
  rho <- matrix(data = 2700,   # density, [kg/(m^3)] (2800)
                nrow = z.num,
                ncol = x.num)
  Cp <- matrix(data = 1000,    # heat capacity, [J/(kg*K)] (1000)
               nrow = z.num,
               ncol = x.num)
  Kappa <- k / (rho * Cp)      # [(m^2)/s]
  Kappa.min <- min(Kappa)      # [(m^2)/s] # Kappa.min <- 1e-6   # (Buchanan)
  U <- matrix(data = 0,        # uplift, [m/s]
              nrow = z.num,
              ncol = x.num)
  # H <- matrix(data = 0,        # latent heat production, value? units?
  #             nrow = z.num,
  #             ncol = x.num)  
  # Input geothermal gradient in K/km.
  a.Kkm <- 13  # [K/km]
  a <- a.Kkm / 1000  # [K/m]
                   
# Define time step.
  dt.max <- (dx ^ 2) / (6 * Kappa.min)  # [s] 
  dt.yr <- dt.max * 3.17098e-8 # [yr]
  dt <- dt.max
  
# Set HL (until it can be properly put into the equation)
  HL <- 0
```

```{r InitializeTDistribution_plume}
# Set up an initial condition of a circular plume at an elevated temperature.
  
# Initialize a matrix for initial temperature conditions.
  T.0 <- matrix(data = 0,
                nrow = z.num,
                ncol = x.num)

# Set the top row to be air temperature.
  # Choose a value for the air temperature.
  T.surface <- 25  # [°C]
  # Assign the first row of the initial matrix to be the air temperature.
  T.0[1, ] <- T.surface + 273.15

# Make a geotherm for the initialization matrix.
  for (j in 2:dim(T.0)[1]) {
    T.0[j, ] <- T.0[1, 1] + a * x[j]
  }
  
# Instantanteously emplace a melt sheet and set material properties of the melt
# sheet.
  # Set dimensions of melt sheet.
  meltsheetdiameter <- 20000  # [m]
  meltsheetthickness <- 1000  # [m]
  # If the melt sheet is centered in the x direction in the model, find the 
  # position in meters along the x axis.
  xstart.melt <- mean(x) - (meltsheetdiameter / 2)  # x position, [m]
  xend.melt <- mean(x) + (meltsheetdiameter / 2)  # x position, [m]
  # Find the index in the vector that corresponds to the position above.
  xstart.melt.index <- which(abs(x - xstart.melt) == min(abs(x - xstart.melt)))
  xend.melt.index <- which(abs(x - xend.melt) == min(abs(x - xend.melt)))
  # Make the melt sheet start on the second row of the matrix, below the air
  # layer.
  zstart.melt.index <- 2 
  # If the melt sheet is meltsheetthickness and starts at the second row of the
  # model, find the index that corresponds to the bottom of the melt sheet.
  zend.melt <- z[2] + meltsheetthickness
  zend.melt.index <- which(abs(z - zend.melt) == min(abs(z - zend.melt)))
  
  # Set temperature of melt sheet.
  T.meltsheet <- 1700  # [°C]
  
  # Emplace melt sheet. Set material properties of melt sheet.
  for (i in zstart.melt.index:zend.melt.index) {
    for (j in xstart.melt.index:xend.melt.index) {
      T.0[i, j] <- T.meltsheet + 273.15
      # k[i, j]     <- 3.0    # thermal conductivity, [W/(m*K)]
      # A[i, j]     <- 1e-6   # heat production, [W/(m^3)]
      # rho[i, j]   <- 3300   # density, [kg/(m^3)]
      # Cp[i, j]    <- 1293   # heat capacity, [J/(kg*K)]
      # Kappa[i, j] <- k[i, j] / (rho[i, j] * Cp[i, j])  # [(m^2)/s]
    }
  }

# Clean up.
  rm(i, j)
```

### Initial Condtions  
  
**Model Parameters:**  
dt = `r round((dt * 3.17098e-8 / 1000000), digits = 3)` Myr  
dx = `r round(dx, digits = 1)` m  
dz = `r round(dz, digits = 1)` m  

**Initial perterbation:**  
rectangular intrusion with `r (meltsheetdiameter / 2 / 1000)` km radius and `r (meltsheetthickness / 1000)` km thickness at an elevated temperature of `r T.meltsheet`°C intrudes into rock with a geothermal gradient of `r a.Kkm` °C/km

```{r PlotInitialConditions_heatmap}
# Plot initial conditions.

# Load packages plotly and Rcolorbrewer for plotting.
  library(plotly)
  library(RColorBrewer)

# Make function for plotting.
  PlotlyMeltSheet <- function(temperatures) {
    # Make matrix for hovertext labels.
    text.matrix <- round(matrix(unlist(temperatures), 
                                ncol = dim(temperatures)[2], 
                                byrow = FALSE) - 273.15, 
                         digits = 0)

    # Make plot.
    plot_ly(x = ~(x / 1000),
            y = ~(z / 1000), 
            z = ~(temperatures - 273.15),
            type = "heatmap",
            colors = rev(brewer.pal(9,"RdYlBu")),
            hoverinfo = "text",
            text = text.matrix) %>%
      add_trace(y = ((z[[2]] - z[[1]])/ 2 / 1000),
                type = "scatter",
                mode = "lines",
                line = list(color = "black")) %>%
      colorbar(title = "Temperature (°C)",
               limits = c(0, 2000))  %>%
      layout(xaxis = list(title = "Distance (km)"),
             yaxis = list(title = "Depth (km)",
                          autorange = "reversed"))
  }

# Plot initial condtions as a heatmap.
  PlotlyMeltSheet(T.0)
```

```{r Model_SetIterations}
# Make a function to solve the finite difference equation for n time steps.
  RunModel <- function(n.timesteps) {
    
    # Initialize a list as a container for model returns at different times.
    T.n  <- list()
    
    # Place T.0 in the first position of the T.n list.
    T.n[[1]] <- T.0
    
    for (t in 1:n.timesteps) {
      
      # Initialize T.n[[t + 1]]
      T.n[[t + 1]] <- matrix(data = NA,
                             nrow = z.num,
                             ncol = x.num)
      
      # Calculate T for interior of T.n[[t + 1]] matrix.
      for (i in 2:(x.num - 1)) {
        for (j in 2:(z.num - 1)) {
          
          # Set constant T border.
          T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
          T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
          T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
          T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
          
          # Calculate T in the interior of the model
          T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
            
            # conduction term
            ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
            
            # heat production term
            ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +
            
            # advection/uplift term
            (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz)) +
            
            # latent heat term
            HL
        }
      }
    }
    return(T.n)
  }
  
# Solve the finite difference equation for n time steps.
  n <- 3
  T.n <- RunModel(n)
```

### Results after `r eval(n)` iterations

Maximum temperature change between model run n = `r n` (time = `r round((n * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap}
# Plot model results after n iterations.
  PlotlyMeltSheet(T.n[[length(T.n)]])
```

```{r Setup_m_iterations}
# Solve the finite difference equation for m time steps.
  m <- 5
  T.m <- RunModel(m)
```

### Results after `r eval(m)` iterations

Maximum temperature change between model run n = `r m` (time = `r round((m * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.m[[length(T.m)]] - T.m[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_m}
# Plot model results after m iterations.
  PlotlyMeltSheet(T.m[[length(T.m)]])
```


```{r Setup_o_iterations}
# Solve the finite difference equation for m time steps.
  o <- 9
  T.o <- RunModel(o)
```

### Results after `r eval(o)` iterations

Maximum temperature change between model run n = `r o` (time = `r round((o * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.o[[length(T.o)]] - T.o[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_o}
# Plot model results after m iterations.
  PlotlyMeltSheet(T.o[[length(T.o)]])
```